A radiocarbon data based evaluation of Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe

### 2019-01-21 article/01_Introduction.Rmd
some discussion text: cultural distance
spelling

How can major misconceptions like this arise and persist in European archaeology? Beyond missing data due to preservation and research history, an additional reason might be a lack of unified data repositories and universally acknowledged methods to measure the spatial and temporal distribution of cultural phenomena. Instead of directly querying archives with quantitative information, the typical process for the creation of large-scale syntheses is still to manually excerpt the literature of all relevant, local research traditions. The present article, in contrast, explores the former strategy. The first major question therefore is: Does openly available, bulk radiocarbon data allow reconstructing the main trends within the distribution of burial rites in Bronze Age Europe? This approach could set a precedent for a more reproducible and statistically sound research practice. If so, then the quantitative comparison of macroregional trends could also pave the way for an analysis of interaction intensity on a long-term continental scale: Do time series constructed from radiocarbon data enable meaningful observations concerning cultural and spatial distance?

@@

Archaeology has a unique perspective on Cultural evolution [@garveyCurrentPotentialRoles2018]: It is concerned with human behaviour over big areas and long timescales. It also works with solid proxies of cultural expression: material remains -- individual lost or discarded artefacts, the ruins of past architecture and all more inconspicuous by-products of living. We rightfully expect humans to learn handcraft as well as building techniques and traditions  -- both functionally meaningful and seemingly pointless -- from others close to them. The similarity of historically and spatially close material remains can, therefore, be explained as a direct result of social interaction [@lymanChronometersUnitsEarly2000; @lymanCultureHistoricalBiological2001; @lymanRiseFallCulture1997; @obrienEpistemologicalNatureArchaeological2002]. Questions of Cultural Transmission form the core of archaeology.

@@

In the following sections I will (1) explain how radiocarbon data with relevant meta information can be transformed to macro regional time series of burial traditions, (2) explore how these timeseries compare to traditional archaeological narratives and (3) construct and contextualise cultural distances between macro regions over time in relation to spatial distance. For the latter a simulation based on the @neimanStylisticVariationEvolutionary1995 template was employed.

>>>>

How can major misconceptions like this arise and persist in European archaeology? Beyond missing data due to preservation and research history, an additional reason might be a lack of unified data repositories and universally acknowledged methods to measure the spatial and temporal distribution of cultural phenomena. Instead of directly querying archives with quantitative information, the typical process for the creation of large-scale syntheses is still to manually excerpt the literature of all relevant, local research traditions. The present article, in contrast, explores the former strategy. The first major question therefore is: Does openly available, bulk radiocarbon data allow reconstructing the main trends within the distribution of burial rites in Bronze Age Europe? This approach could set a precedent for a more reproducible and statistically sound research practice. If so, then the quantitative comparison of macro regional trends could also pave the way for an analysis of interaction intensity on a long-term continental scale: Do time series constructed from radiocarbon data enable meaningful observations concerning cultural and spatial distance?

@@

Archaeology has a unique perspective on Cultural evolution [@garveyCurrentPotentialRoles2018]: It is concerned with human behaviour over big areas and long time scales. It also works with solid proxies of cultural expression: material remains -- individual lost or discarded artefacts, the ruins of past architecture and all more inconspicuous by-products of living. We rightfully expect humans to learn handcraft as well as building techniques and traditions  -- both functionally meaningful and seemingly pointless -- from others close to them. The similarity of historically and spatially close material remains can, therefore, be explained as a direct result of social interaction [@lymanChronometersUnitsEarly2000; @lymanCultureHistoricalBiological2001; @lymanRiseFallCulture1997; @obrienEpistemologicalNatureArchaeological2002]. Questions of Cultural Transmission form the core of archaeology.

@@

In the following sections I will (1) explain how radiocarbon data with relevant meta information can be transformed to macro regional time series of burial traditions, (2) explore how these time series compare to traditional archaeological narratives and (3) construct and contextualise cultural distances between macro regions over time in relation to spatial distance. For the latter a simulation based on the @neimanStylisticVariationEvolutionary1995 template was employed.

### 2019-01-21 article/02_Materials_and_Methods.Rmd
spelling

These time series also allow to derive another set of time series with a measure of cultural distance for each region-region relationship (Figure \ref{sed_region_matrix}). Cultural distance is defined here as the squared euclidian distance (SED) between the trait proportions of two different regions at a certain point in time [@neimanStylisticVariationEvolutionary1995]. SED seems to be a sensible statistic here, because it is relatively simple, works with only a few assumptions and emphasises differences through squaring.

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../analysis/figures/sed_region_matrix.jpeg}
\caption{\label{sed_region_matrix}Plot matrix of the region-region squared euclidian distance time series. In the top right half for the variable \textit{burial type} and in the bottom left for the variable \textit{burial construction}. On the x-axis the time is displayed, on the y-axis the squared euclidian distance between the two relevant regions. The faint, grey line shows the actual data, the stronger, coloured line an approximation (LOESS with $span = 0.3$).}
\end{figure*}

@@

To test for correlation between the distances based on the variables *burial type* and *burial construction*, the timeframe of interest was divided into sections of 200 years. For each section of each the two variables, a matrix of the mean SED over time was calculated. The two matrices of each section can be tested for correlation with the Mantel test, which uses permutations of allegedly correlating matrices to determine how strongly random reordering can alter an underlying correlation coefficient [@mantelDetectionDiseaseClustering1967]. Since both distances compared here, *burial type* and *burial construction* distance, are cardinal scale variables, the Pearson correlation coefficient [@pearsonNotesRegressionInheritance1895] can be applied as a base measure for the Mantel test (Table \ref{tab:distanceandcorrelationtable*} C).

@@

To get a better understanding how the correlation coefficient should develop if there was a strong correlation between spatial and cultural distance as measured by *burial type* and *burial construction*, a simple simulation was implemented based on the model proposed by @neimanStylisticVariationEvolutionary1995. This simulation assumes a population separated into $g = 8$ groups (regions) with $N_g = 100$ individuals each. $k = 2$ mutually exclusive ideas are randomly but equally distributed in this population at $t_{0}$. With every timestep ($t_{total} = (2200 - 800)/20 = 70$) all individuals learn from another randomly chosen individual. Several different values where tested for the probability $m_i$ that the other agent the individual is learning from comes not from the same, but from another group:

@@

The partner group for each learning instance is selected randomly following a group interaction probability matrix $I$. Two different matrizes where tested:

@@

The distribution of the $k$ variants within the groups can be observed for every timestep and documented in time series just like the ones for the real-world regions data (Figure \ref{development_burial_type} B and \ref{development_burial_construction} B). These time series can as well be used to calculate time series of region-region distance, which again can be split up into time blocks of 200 years to calculate mean distance matrices. The latter can be checked for correlation with the spatial distance matrix (cf. Table \ref{tab:distanceandcorrelationtable*} AB) and compared with other simulation results and real-world observations (Figure \ref{simulation}).

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../analysis/figures/simulation.jpeg}
\caption{\label{simulation}Development of correlation between cultural and spatial distance for real world data and simulation results. The plots in the left column of the plot matrix (\textbf{ACE}) contain simulation results for an equal group distance matrix $I$, the ones on the right (\textbf{BDF}) for a distance matrix based on the macro unit distances defined for this paper (Figure \ref{map_regions} B). The rows of the plot matrix (\textbf{AB}, \textbf{CD}, \textbf{EF}) represent different degrees of intergroup transmission $m_i$. Each plot shows the results of 50 simulation runs. The x-axis displays the time distiguished into 200 year time slots, the y-axis Spearman's rank correlation coefficient between the mean squared euclidian distance matrix for a 200 year time slot and the static spatial distance matrix. This information is visualized for each time slot with (from left to right) a dotplot coloured according to the $p$-value of the Mantel test, a scatter plot line and a boxplot. The scatter plot points of one simulation run are connected with a faint line from one time slot to the next. To compare the runs with the real world observations, the results for the burial customs data (Table \ref{tab:distanceandcorrelationtable*} AB) is added here with with a stronger green and blue line. The strong horizontal red line marks the correlation coefficient of zero, where neither positive nor negative correlation can be assumed.}
\end{figure*}

>>>>

These time series also allow to derive another set of time series with a measure of cultural distance for each region-region relationship (Figure \ref{sed_region_matrix}). Cultural distance is defined here as the squared Euclidean distance (SED) between the trait proportions of two different regions at a certain point in time [@neimanStylisticVariationEvolutionary1995]. SED seems to be a sensible statistic here, because it is relatively simple, works with only a few assumptions and emphasises differences through squaring.

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../analysis/figures/sed_region_matrix.jpeg}
\caption{\label{sed_region_matrix}Plot matrix of the region-region squared Euclidean distance time series. In the top right half for the variable \textit{burial type} and in the bottom left for the variable \textit{burial construction}. On the x-axis the time is displayed, on the y-axis the squared Euclidean distance between the two relevant regions. The faint, grey line shows the actual data, the stronger, coloured line an approximation (LOESS with $span = 0.3$).}
\end{figure*}

@@

To test for correlation between the distances based on the variables *burial type* and *burial construction*, the time frame of interest was divided into sections of 200 years. For each section of each the two variables, a matrix of the mean SED over time was calculated. The two matrices of each section can be tested for correlation with the Mantel test, which uses permutations of allegedly correlating matrices to determine how strongly random reordering can alter an underlying correlation coefficient [@mantelDetectionDiseaseClustering1967]. Since both distances compared here, *burial type* and *burial construction* distance, are cardinal scale variables, the Pearson correlation coefficient [@pearsonNotesRegressionInheritance1895] can be applied as a base measure for the Mantel test (Table \ref{tab:distanceandcorrelationtable*} C).

@@

To get a better understanding how the correlation coefficient should develop if there was a strong correlation between spatial and cultural distance as measured by *burial type* and *burial construction*, a simple simulation was implemented based on the model proposed by @neimanStylisticVariationEvolutionary1995. This simulation assumes a population separated into $g = 8$ groups (regions) with $N_g = 100$ individuals each. $k = 2$ mutually exclusive ideas are randomly but equally distributed in this population at $t_{0}$. With every time step ($t_{total} = (2200 - 800)/20 = 70$) all individuals learn from another randomly chosen individual. Several different values where tested for the probability $m_i$ that the other agent the individual is learning from comes not from the same, but from another group:

@@

The partner group for each learning instance is selected randomly following a group interaction probability matrix $I$. Two different matrices where tested:

@@

The distribution of the $k$ variants within the groups can be observed for every time step and documented in time series just like the ones for the real-world regions data (Figure \ref{development_burial_type} B and \ref{development_burial_construction} B). These time series can as well be used to calculate time series of region-region distance, which again can be split up into time blocks of 200 years to calculate mean distance matrices. The latter can be checked for correlation with the spatial distance matrix (cf. Table \ref{tab:distanceandcorrelationtable*} AB) and compared with other simulation results and real-world observations (Figure \ref{simulation}).

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{../analysis/figures/simulation.jpeg}
\caption{\label{simulation}Development of correlation between cultural and spatial distance for real world data and simulation results. The plots in the left column of the plot matrix (\textbf{ACE}) contain simulation results for an equal group distance matrix $I$, the ones on the right (\textbf{BDF}) for a distance matrix based on the macro unit distances defined for this paper (Figure \ref{map_regions} B). The rows of the plot matrix (\textbf{AB}, \textbf{CD}, \textbf{EF}) represent different degrees of intergroup transmission $m_i$. Each plot shows the results of 50 simulation runs. The x-axis displays the time distinguished into 200 year time slots, the y-axis Spearman's rank correlation coefficient between the mean squared Euclidean distance matrix for a 200 year time slot and the static spatial distance matrix. This information is visualized for each time slot with (from left to right) a dotplot coloured according to the $p$-value of the Mantel test, a scatter plot line and a boxplot. The scatter plot points of one simulation run are connected with a faint line from one time slot to the next. To compare the runs with the real world observations, the results for the burial customs data (Table \ref{tab:distanceandcorrelationtable*} AB) is added here with with a stronger green and blue line. The strong horizontal red line marks the correlation coefficient of zero, where neither positive nor negative correlation can be assumed.}
\end{figure*}

### 2019-01-21 article/03_Results.Rmd
spelling

In the course of the Bronze Age in Europe there was a trend away from body burial towards cremation. Around 2200calBC cremation burials were almost unknown in Poland, southern Germany, northeastern France and northern Germany. In the northwest, in Great Britain and the Benelux area, and in the southeast, in Austria and the Czech Republic, however, there were early cremation traditions. These contexts could be discussed as areas of origin of the later omnipresent phenomenon. While cremation burials lost importance in Austria and the Czech Republic as well as in the Benelux region during the Early Bronze Age and only regained their importance in the Middle Bronze Age, their share increased steadily in England. Inhumation and cremation were balanced here for a long time. The situation was similar in southern Scandinavia, where the proportion of cremation burials was much lower until the Late Bronze Age. In Poland and northern Germany, cremation became relatively suddenly and completely established with the beginning of the Late Bronze Age. In northeastern France and northern Germany this change took place already in the Middle Bronze Age. In both contexts, inhumation continued to play an important role.

@@

Beyond the simple comparison of archaeological knowledge and information derived from Radon-B, Figure \ref{sed_region_matrix} shows the time series of squared euclidian distance for each region-region relationship. The majority of the curves in its top right half for *burial type* correspond to expectations: At the beginning of the Bronze Age, inhumation is widespread throughout the study area, so the Euclidean distance at 2200calBC is small in almost every region-region relationship. Only the Benelux region, where the share of cremation is already high in the Early Bronze Age, is excluded. From the Middle Bronze Age onwards, but especially at the transition to the Late Bronze Age, the transition to the dominance of cremation took place all over Europe. The SED values increase in this period and show clear peaks or several successive maxima. This rise is a consequence of the time lag in the transition processes as depicted in Radon-B: It is different in every region. At the end of the Late Bronze Age, the average SED is approaching zero again, as burial by cremation dominates in almost all of Europe. The early and complete adoption of the cremation rite in the Benelux area, as depicted in Radon-B, clearly sets it apart from the other regions. England, on the other hand, with its slow, gradual transition from inhumation to cremation, stands out due to its consistently moderate distances to all other regions. It remains without accentuated minima and maxima. The situation for *burial construction* is more complicated, as can be seen from the plots in the bottom right corner of Figure \ref{sed_region_matrix}. The distance does not seem to follow a general trend, but a multitude of local and temporally limited phenomena. This heterogeneity is of course also due to how correctly or incorrectly the proxy developed from Radon-B depicts developments in the real world. In the Middle Bronze Age, the Tumulus culture should bring about a fundamental reduction of the SED for almost all region-region relationships -- this effect is indeed weakly visible.

>>>>

In the course of the Bronze Age in Europe there was a trend away from body burial towards cremation. Around 2200calBC cremation burials were almost unknown in Poland, southern Germany, northeastern France and northern Germany. In the Northwest, in Great Britain and the Benelux area, and in the Southeast, in Austria and the Czech Republic, however, there were early cremation traditions. These contexts could be discussed as areas of origin of the later omnipresent phenomenon. While cremation burials lost importance in Austria and the Czech Republic as well as in the Benelux region during the Early Bronze Age and only regained their importance in the Middle Bronze Age, their share increased steadily in England. Inhumation and cremation were balanced here for a long time. The situation was similar in southern Scandinavia, where the proportion of cremation burials was much lower until the Late Bronze Age. In Poland and northern Germany, cremation became relatively suddenly and completely established with the beginning of the Late Bronze Age. In northeastern France and northern Germany this change took place already in the Middle Bronze Age. In both contexts, inhumation continued to play an important role.

@@

Beyond the simple comparison of archaeological knowledge and information derived from Radon-B, Figure \ref{sed_region_matrix} shows the time series of squared Euclidean distance for each region-region relationship. The majority of the curves in its top right half for *burial type* correspond to expectations: At the beginning of the Bronze Age, inhumation is widespread throughout the study area, so the Euclidean distance at 2200calBC is small in almost every region-region relationship. Only the Benelux region, where the share of cremation is already high in the Early Bronze Age, is excluded. From the Middle Bronze Age onwards, but especially at the transition to the Late Bronze Age, the transition to the dominance of cremation took place all over Europe. The SED values increase in this period and show clear peaks or several successive maxima. This rise is a consequence of the time lag in the transition processes as depicted in Radon-B: It is different in every region. At the end of the Late Bronze Age, the average SED is approaching zero again, as burial by cremation dominates in almost all of Europe. The early and complete adoption of the cremation rite in the Benelux area, as depicted in Radon-B, clearly sets it apart from the other regions. England, on the other hand, with its slow, gradual transition from inhumation to cremation, stands out due to its consistently moderate distances to all other regions. It remains without accentuated minima and maxima. The situation for *burial construction* is more complicated, as can be seen from the plots in the bottom right corner of Figure \ref{sed_region_matrix}. The distance does not seem to follow a general trend, but a multitude of local and temporally limited phenomena. This heterogeneity is of course also due to how correctly or incorrectly the proxy developed from Radon-B depicts developments in the real world. In the Middle Bronze Age, the Tumulus culture should bring about a fundamental reduction of the SED for almost all region-region relationships -- this effect is indeed weakly visible.

### 2019-01-21 article/04_Discussion.Rmd
spelling

One major question asked in this paper was, if openly available, bulk radiocarbon data does allow the reconstruction of main trends within the distribution of burial rites in Bronze Age Europe. This question can be answered with Yes. The narrative in section \ref{time-series-results} is surprisingly similar to the one presented by @hardingEuropeanSocietiesBronze2000a. The following paragraphs contain a short evaluation of the time series in Figure \ref{development_burial_type} and \ref{development_burial_construction} in the light of the archaeological state of research for each region. Despite the general success of this methodological apporach, unfortunately the amount and quality of data is still not sufficient for a comprehensive picture. Please see the discussion of data quality in section \ref{data-quality-discussion}. \newline

`r a("Southeastern Central Europe")`^[Amount of graves in Radon-B for this region.]:
The artificial circular region *Southeastern Central Europe* comprises the eastern part of Austria (Upper Austria, Lower Austria, Styria, Burgenland), the northwest of Hungary, western Slovakia, Moravia and large parts of Bohemia (Figure \ref{map_regions} A). The density of ^14^C data on graves in Radon-B is low -- dates are available from the various regions, but large areas are without evidence. Diachronically the amount of data is consistently low and reaches a minimum in the Middle Bronze Age. While there is a constant flow of information for the variable *burial type*, the number of graves without information on *burial construction* increases in the Late Bronze Age to such an extent that no meaningful overview can be derived. The region intersects diverse geographical and cultural areas complicating the comparison with archaeological observations. In the southeast of Central Europe, cremation and inhumation already existed side by side in the Early Bronze Age, and the local tradition remained stable in this respect also due to the expansion of the burial mound tradition in the Middle Bronze Age -- with a slight overall increase in cremation graves. The Urnfield period brought the transition to the absolute dominance of cremation burial. The development of the *burial type* variable in the Radon-B proxy shows this development correctly. The information from ^14^C data on *burial construction* correctly traces the relevance of flat burials in the Early Bronze Age but then overemphasises them in the Middle Bronze Age compared to the construction of burial mounds. In the Late Bronze Age, burials on flat burial grounds were frequent, although burial mounds of the Tumulus Culture were also further documented. This impact does not become clear in the Radon-B data -- indeed due to the small number of dates [@lubosCzechLandsAustria2013; @markovaSlovakiaHungary2013].

@@

`r a("Southern Germany")`:
The artificial district region *Southern Germany* covers Bavaria completely but also includes most of Baden-Württemberg and Thuringia as well as extensive areas in Moravia and western Austria. No ^14^C data are available from the latter regions, the majority of the available data comes from Upper Palatinate, Lower Bavaria and Swabia as well as (in Württemberg) the administrative district of Tübingen. For the Early Bronze Age, Radon-B preserves a remarkable amount of data from southern Germany, but this amount decreases in the course of the Middle Bronze Age and leaves the Late Bronze Age almost without dated graves. The *burial type* variable is quite well understood -- although disproportionately much more of the early inhumations were recorded than urnfield burials. This wealth of information contrasts sharply with the situation regarding the *burial construction* variable, for which almost no information from this area is available throughout the Bronze Age. If one considers the relative development calculated from Radon-B and compares it with the archaeological literature, then a reasonable agreement can be found for *burial type*. Inhumation dominated for a long time but was then replaced by cremation from the Middle Bronze Age onwards and massively in the Late Bronze Age. The calculated proportions for *burial construction* are based on a negligibly small amount of data, but the proxy still indicates the right tendency for early and middle Bronze Age: Flat graves replaced mounds. The dataset does not allow any statements about the Late Bronze Age [@falkensteinWandelBestattungssittenHugelgraber2017; @jockenhovelGermanyBronzeAge2013].

@@

`r a("Southern Scandinavia")`:
The region *Southern Scandinavia* includes all of Denmark with Jutland, Funen, Zeeland and all smaller islands as well as large parts of east of southern Sweden, and coastal regions in Schleswig-Holstein and Mecklenburg-Western Pomerania. The ^14^C data are very well distributed across Jutland, Funen and the Swedish Skåne region. Since the Bronze Age in Scandinavia begins much later than in Central Europe and is measured by another chronological system, the diachronic developments cannot merely be compared through the tripartite structure of the Early, Middle and Late Bronze Ages. The period which is labelled Early Bronze Age in Central Europe corresponds to the Scandinavian Late Neolithic. The majority of data from Scandinavia are from the period IB onwards. In this respect, the data density for almost the entire Bronze Age is good -- also with regard to the two variables *burial type* and *burial construction*. The Late Neolithic from 2200calBC is a desideratum. If one carries out the necessary mental shift -- the Central European Middle Bronze Age corresponds to the Nordic Early Bronze Age -- and compares the proxy of the relative grave form development with the archaeological observations, then one finds a substantial agreement: inhumations were the rule in periods I and II (about 1800-1300calBC), but cremations also occurred sporadically. From period III onwards cremation burials gained the upper hand. In the Scandinavian Early Bronze Age burials in burial mounds dominated. In the Late Bronze Age, flat burial fields began to occure [@thraneScandinavia2013].

@@

`r a("England")`:
The *England* region stretches across England up to York and Liverpool, incorporating parts of Wales and leaving out Cornwall and Devon in the southwest. It also touches a narrow coastal strip on the mainland, from which no ^14^C data are documented. The available data are concentrated in the southeast of Britain but are well distributed there. The diachronic data availability is relatively good in the early and middle Bronze Age but then decreases in the late Bronze Age. *burial type* and *burial construction* behave accordingly. The relative development derived from the ^14^C data indicates the correct trend for *burial type*: Cremation graves became more and more important during the Bronze Age, while inhumations appeared for a long time in parallel. However, the dataset emphasises the occurence of cremation graves too much. Concerning the variables *burial construction*, the dominance of burial mounds corresponds perfectly to the archaeological narrative, but -- especially in the Late Bronze Age -- flat grave fields also occurred, which the data set wrongly excludes [@robertsBritainIrelandBronze2013].

@@

The expectation for this question was a clear correlation of the "cultural" SED distances for both variables *burial type* and *burial construction* as well in between these variables and the spatial distance. The simulation, though with this implementation and parametrisation for sure rather qualitativly than quantitativly, visualises how a high degree of interaction should translate to a significantly higher degree of cultural similarity across human populations (Figure \ref{simulation}). And as the European Bronze Age is usually depicted as a period of intensive trade, travel and exchange [e.g. @kristiansenRiseBronzeAge2005; @sherrattWhatWouldBronzeAge1993] there was no reason to expect something else than Tobler's famous First Law of Geography: "everything is related to everything else, but near things are more related than distant things." [@toblerComputerMovieSimulating1970].

@@

1. The amount, distribution and quality of the data as discussed in section \ref{data-quality-discussion} queries the credibility of the cultural distance calculation. The macroregions have to be very big due to a lack of data and don't allow for a precise picture.

@@

Keeping this in mind, the surprising result is still valid and has to be explained. One interpretation might be, that the Early Bronze Age was a time of rising cultural exchange and religious homogenization, while in later periods different domains secluded themselves more strongly. The negative correlation of *burial type* in the Late Bronze Age could indicate long distance contacts that explicitly excluded certain intermediate regions of Central Europe. The rise of the Tumulus Culture in the Middle Bronze Age might have cushioned this effect for the variable *burial construction* and at the same time introduced more variability for a less important aspect *burial type*. The latter may have been of subordinate relevance in the Tumulus Culture context. With the rise of the Urnfield Culture this phenomonon could have been reversed as *burial type* became an expression of something that could have been one of the first formalized religions in Europe: The spread of the Urnfield Culture was an extraordinary development in early European history and was possibly accompanied by a structured, religious dogma that could have built up around the symbolic content of cult wagons, sunbarks and waterfowl [@bilicSwanChariotSolar2016; @kimmig_seevolkerbewegung_1964; @molloy_european_2018]. Of course these explanations could, if accurate, only be considered on a really big scale. For individual regions and sites specific ideological, social or even economic considerations might have been much more critical for the adopted burial tradition [@barrett_monumentality_1990; @fokkens_genesis_1997-1; @muller-scheesel_variabilitat_2009].

@@

3. In the field of investigation of this work different research traditions meet, which in the past considered ^14^C data more or less essential for the construction of an absolute chronological assessment. In various Metal Age contexts, 14C data are inferior to the accuracy of relative chronological classifications - e.g. by primer typology - and are therefore used only sporadically to create absolute chronological fixpoints. Dendrochronological dating is preferred in moist soil contexts and when the necessary woods are available. The available ^14^C calibration curves form plateaus in some periods, which significantly limit the possible dating accuracy of the data. If such an effect is known, then no data is ordered at all.

4. ^14^C dating requires a small but sufficiently contextualizable amount of organic finds. Due to the old wood and old water effect, short-lived sample materials such as bones of terrestrial organisms or charred cereal grains are preferred. The majority of ^14^C data for this work comes from bones and teeth. However, their preservation is subject to huge differences depending on the soil [@brockPrescreeningTechniquesIdentification2010; @gordonSoilPHBone1981; @nielsen-marshBoneDiagenesisEuropean2007]. In many mineral soil contexts no sufficiently good, dateable material is available. This situation can affect entire regions, if, for example, poor bone preservation prevails in general as a result of low-calcification soils.

>>>>

One major question asked in this paper was, if openly available, bulk radiocarbon data does allow the reconstruction of main trends within the distribution of burial rites in Bronze Age Europe. This question can be answered with Yes. The narrative in section \ref{time-series-results} is surprisingly similar to the one presented by @hardingEuropeanSocietiesBronze2000a. The following paragraphs contain a short evaluation of the time series in Figure \ref{development_burial_type} and \ref{development_burial_construction} in the light of the archaeological state of research for each region. Despite the general success of this methodological approach, unfortunately the amount and quality of data is still not sufficient for a comprehensive picture. Please see the discussion of data quality in section \ref{data-quality-discussion}. \newline

`r a("Southeastern Central Europe")`^[Amount of graves in Radon-B for this region.]:
The artificial circular region *Southeastern Central Europe* comprises the eastern part of Austria (Upper Austria, Lower Austria, Styria, Burgenland), the northwest of Hungary, western Slovakia, Moravia and large parts of Bohemia (Figure \ref{map_regions} A). The density of ^14^C data on graves in Radon-B is low -- dates are available from the various regions, but large areas are without evidence. Diachronically the amount of data is consistently low and reaches a minimum in the Middle Bronze Age. While there is a constant flow of information for the variable *burial type*, the number of graves without information on *burial construction* increases in the Late Bronze Age to such an extent that no meaningful overview can be derived. The region intersects diverse geographical and cultural areas complicating the comparison with archaeological observations. In the Southeast of Central Europe, cremation and inhumation already existed side by side in the Early Bronze Age, and the local tradition remained stable in this respect also due to the expansion of the burial mound tradition in the Middle Bronze Age -- with a slight overall increase in cremation graves. The Urnfield period brought the transition to the absolute dominance of cremation burial. The development of the *burial type* variable in the Radon-B proxy shows this development correctly. The information from ^14^C data on *burial construction* correctly traces the relevance of flat burials in the Early Bronze Age but then overemphasises them in the Middle Bronze Age compared to the construction of burial mounds. In the Late Bronze Age, burials on flat burial grounds were frequent, although burial mounds of the Tumulus Culture were also further documented. This impact does not become clear in the Radon-B data -- indeed due to the small number of dates [@lubosCzechLandsAustria2013; @markovaSlovakiaHungary2013].

@@

`r a("Southern Germany")`:
The artificial district region *Southern Germany* covers Bavaria completely but also includes most of Baden-Württemberg and Thuringia as well as extensive areas in Moravia and western Austria. No ^14^C data are available from the latter regions, the majority of the available data comes from Upper Palatinate, Lower Bavaria and Swabia as well as (in Württemberg) the administrative district of Tübingen. For the Early Bronze Age, Radon-B preserves a remarkable amount of data from southern Germany, but this amount decreases in the course of the Middle Bronze Age and leaves the Late Bronze Age almost without dated graves. The *burial type* variable is quite well understood -- although disproportionately much more of the early inhumations were recorded than Urnfield burials. This wealth of information contrasts sharply with the situation regarding the *burial construction* variable, for which almost no information from this area is available throughout the Bronze Age. If one considers the relative development calculated from Radon-B and compares it with the archaeological literature, then a reasonable agreement can be found for *burial type*. Inhumation dominated for a long time but was then replaced by cremation from the Middle Bronze Age onwards and massively in the Late Bronze Age. The calculated proportions for *burial construction* are based on a negligibly small amount of data, but the proxy still indicates the right tendency for early and middle Bronze Age: Flat graves replaced mounds. The dataset does not allow any statements about the Late Bronze Age [@falkensteinWandelBestattungssittenHugelgraber2017; @jockenhovelGermanyBronzeAge2013].

@@

`r a("Southern Scandinavia")`:
The region *Southern Scandinavia* includes all of Denmark with Jutland, Funen, Zeeland and all smaller islands as well as large parts of east of southern Sweden, and coastal regions in Schleswig-Holstein and Mecklenburg-Western Pomerania. The ^14^C data are very well distributed across Jutland, Funen and the Swedish Skåne region. Since the Bronze Age in Scandinavia begins much later than in Central Europe and is measured by another chronological system, the diachronic developments cannot merely be compared through the tripartite structure of the Early, Middle and Late Bronze Ages. The period which is labelled Early Bronze Age in Central Europe corresponds to the Scandinavian Late Neolithic. The majority of data from Scandinavia are from the period IB onwards. In this respect, the data density for almost the entire Bronze Age is good -- also with regard to the two variables *burial type* and *burial construction*. The Late Neolithic from 2200calBC is a desideratum. If one carries out the necessary mental shift -- the Central European Middle Bronze Age corresponds to the Nordic Early Bronze Age -- and compares the proxy of the relative grave form development with the archaeological observations, then one finds a substantial agreement: inhumations were the rule in periods I and II (about 1800-1300calBC), but cremations also occurred sporadically. From period III onwards cremation burials gained the upper hand. In the Scandinavian Early Bronze Age burials in burial mounds dominated. In the Late Bronze Age, flat burial fields began to occur [@thraneScandinavia2013].

@@

`r a("England")`:
The *England* region stretches across England up to York and Liverpool, incorporating parts of Wales and leaving out Cornwall and Devon in the Southwest. It also touches a narrow coastal strip on the mainland, from which no ^14^C data are documented. The available data are concentrated in the Southeast of Britain but are well distributed there. The diachronic data availability is relatively good in the early and middle Bronze Age but then decreases in the late Bronze Age. *burial type* and *burial construction* behave accordingly. The relative development derived from the ^14^C data indicates the correct trend for *burial type*: Cremation graves became more and more important during the Bronze Age, while inhumations appeared for a long time in parallel. However, the dataset emphasises the occurrence of cremation graves too much. Concerning the variables *burial construction*, the dominance of burial mounds corresponds perfectly to the archaeological narrative, but -- especially in the Late Bronze Age -- flat grave fields also occurred, which the data set wrongly excludes [@robertsBritainIrelandBronze2013].

@@

The expectation for this question was a clear correlation of the "cultural" SED distances for both variables *burial type* and *burial construction* as well in between these variables and the spatial distance. The simulation, though with this implementation and parametrisation for sure rather qualitatively than quantitatively, visualises how a high degree of interaction should translate to a significantly higher degree of cultural similarity across human populations (Figure \ref{simulation}). And as the European Bronze Age is usually depicted as a period of intensive trade, travel and exchange [e.g. @kristiansenRiseBronzeAge2005; @sherrattWhatWouldBronzeAge1993] there was no reason to expect something else than Tobler's famous First Law of Geography: "everything is related to everything else, but near things are more related than distant things." [@toblerComputerMovieSimulating1970].

@@

1. The amount, distribution and quality of the data as discussed in section \ref{data-quality-discussion} queries the credibility of the cultural distance calculation. The macro regions have to be very big due to a lack of data and don't allow for a precise picture.

@@

Keeping this in mind, the surprising result is still valid and has to be explained. One interpretation might be, that the Early Bronze Age was a time of rising cultural exchange and religious homogenization, while in later periods different domains secluded themselves more strongly. The negative correlation of *burial type* in the Late Bronze Age could indicate long distance contacts that explicitly excluded certain intermediate regions of Central Europe. The rise of the Tumulus Culture in the Middle Bronze Age might have cushioned this effect for the variable *burial construction* and at the same time introduced more variability for a less important aspect *burial type*. The latter may have been of subordinate relevance in the Tumulus Culture context. With the rise of the Urnfield Culture this phenomenon could have been reversed as *burial type* became an expression of something that could have been one of the first formalized religions in Europe: The spread of the Urnfield Culture was an extraordinary development in early European history and was possibly accompanied by a structured, religious dogma that could have built up around the symbolic content of cult wagons, sun barks and waterfowl [@bilicSwanChariotSolar2016; @kimmig_seevolkerbewegung_1964; @molloy_european_2018]. Of course these explanations could, if accurate, only be considered on a really big scale. For individual regions and sites specific ideological, social or even economic considerations might have been much more critical for the adopted burial tradition [@barrett_monumentality_1990; @fokkens_genesis_1997-1; @muller-scheesel_variabilitat_2009].

@@

3. In the field of investigation of this work different research traditions meet, which in the past considered ^14^C data more or less essential for the construction of an absolute chronological assessment. In various Metal Age contexts, 14C data are inferior to the accuracy of relative chronological classifications - e.g. by primer typology - and are therefore used only sporadically to create absolute chronological fix points. Dendrochronological dating is preferred in moist soil contexts and when the necessary woods are available. The available ^14^C calibration curves form plateaus in some periods, which significantly limit the possible dating accuracy of the data. If such an effect is known, then no data is ordered at all.

4. ^14^C dating requires a small but sufficiently contextualised amount of organic finds. Due to the old wood and old water effect, short-lived sample materials such as bones of terrestrial organisms or charred cereal grains are preferred. The majority of ^14^C data for this work comes from bones and teeth. However, their preservation is subject to huge differences depending on the soil [@brockPrescreeningTechniquesIdentification2010; @gordonSoilPHBone1981; @nielsen-marshBoneDiagenesisEuropean2007]. In many mineral soil contexts no sufficiently good, dateable material is available. This situation can affect entire regions, if, for example, poor bone preservation prevails in general as a result of low-calcification soils.
